---
title: "First Pitch Equilibrium Analysis"
format:
  html:
    code-fold: true
    theme: cosmo
---

```{r setup, include=FALSE}
# Load required libraries
library(tidyverse)
library(Lahman)
library(scales)

# Suppress startup messages
suppressPackageStartupMessages({
  library(tidyverse)
  library(Lahman)
  library(scales)
})

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```

## User Configuration

```{r user_inputs}
# ============================================================================
# USER INPUTS: Configure your analysis here
# ============================================================================

# DATA SOURCE SELECTION
# Choose which data to use: "lahman" or "statcast"
data_source <- "statcast"  # Options: "lahman", "statcast"

# PLAYER SELECTION (format: "Last, First")
# Set to NULL to use League Average stats
# Examples: "Rodón, Carlos", "Ward, Taylor", "Sale, Chris"

target_batter <- NULL
target_pitcher <- NULL

# Optional: Define manual stats (uncomment and modify to use)
# manual_batter_stats <- c(
#   swing_rate = 0.45,
#   contact_rate = 0.85,
#   zone_rate = 0.48
# )
#
# manual_pitcher_stats <- c(
#   zone_rate = 0.45,
#   chase_rate = 0.30,
#   strike_rate = 0.65
# )
```

## Theme Setup

```{r theme_setup}
# ============================================================================
# CUSTOM BASEBALL THEME FOR VISUALIZATIONS
# ============================================================================

# Baseball color palette
baseball_colors <- list(
  dark_green = "#003831",    # Classic baseball field green
  red = "#BD3039",           # Baseball seam red
  white = "#FFFFFF",         # Baseball white
  cream = "#F5F5DC",         # Vintage baseball cream
  brown = "#8B4513",         # Dirt/leather brown
  blue = "#002D72"           # Classic baseball blue
)

# Custom ggplot theme
theme_baseball <- function() {
  theme_minimal() +
    theme(
      plot.title = element_text(size = 16, face = "bold", color = baseball_colors$dark_green),
      plot.subtitle = element_text(size = 12, color = baseball_colors$brown),
      axis.title = element_text(size = 11, face = "bold"),
      axis.text = element_text(size = 10),
      panel.grid.major = element_line(color = "grey85", linewidth = 0.3),
      panel.grid.minor = element_line(color = "grey90", linewidth = 0.2),
      legend.position = "bottom",
      legend.title = element_text(face = "bold"),
      plot.background = element_rect(fill = baseball_colors$cream, color = NA),
      panel.background = element_rect(fill = "white", color = NA)
    )
}
```

## Load Statcast Data

```{r load_statcast, message=FALSE}
# ============================================================================
# LOAD 2024 STATCAST CSV FILES
# ============================================================================

# Load CSV files if using Statcast data
if (data_source == "statcast") {

  # Load batter stats
  statcast_batters <- read_csv("batter_stats_2024.csv", show_col_types = FALSE) %>%
    rename(player_name = `last_name, first_name`) %>%
    mutate(player_name = str_trim(player_name))

  # Load pitcher stats
  statcast_pitchers <- read_csv("pitcher_stats_2024.csv", show_col_types = FALSE) %>%
    rename(player_name = `last_name, first_name`) %>%
    mutate(player_name = str_trim(player_name))

  cat("✅ Statcast data loaded successfully\n")
  cat("   Batters:", nrow(statcast_batters), "players\n")
  cat("   Pitchers:", nrow(statcast_pitchers), "players\n\n")

} else {
  cat("📚 Using Lahman database (Statcast CSVs not loaded)\n\n")
}
```

## Data Ingestion

```{r data_functions}
# ============================================================================
# DATA INGESTION FUNCTIONS
# ============================================================================

get_batter_stats <- function(name = NULL) {
  # Decision tree for data source selection:
  # 1. If manual_batter_stats exists -> Use manual stats
  # 2. If name is provided -> Fetch from Lahman database
  # 3. If name is NULL or not found -> Calculate League Average for most recent year

  # Check for manual stats first
  if (exists("manual_batter_stats") && !is.null(manual_batter_stats)) {
    cat("📊 Using MANUAL stats for batter\n")
    return(manual_batter_stats)
  }

  # If a specific name is provided, fetch from Lahman
  if (!is.null(name)) {
    # Parse name (format: "Last, First")
    name_parts <- str_split(name, ", ")[[1]]
    if (length(name_parts) == 2) {
      last_name <- name_parts[1]
      first_name <- name_parts[2]

      # Find player in People table
      player <- People %>%
        filter(nameLast == last_name, nameFirst == first_name) %>%
        slice(1)

      if (nrow(player) > 0) {
        player_id <- player$playerID

        # Get most recent season stats
        recent_year <- max(Batting$yearID)
        player_stats <- Batting %>%
          filter(playerID == player_id, yearID == recent_year) %>%
          summarise(
            avg = sum(H, na.rm = TRUE) / sum(AB, na.rm = TRUE),
            obp = sum(H + BB + HBP, na.rm = TRUE) / sum(AB + BB + HBP + SF, na.rm = TRUE),
            slg = sum(H + X2B + 2*X3B + 3*HR, na.rm = TRUE) / sum(AB, na.rm = TRUE),
            pa = sum(AB + BB + HBP + SF, na.rm = TRUE)
          )

        if (nrow(player_stats) > 0 && player_stats$pa > 0) {
          cat("⚾ Using LAHMAN DATABASE stats for batter:", name, "\n")
          return(as.list(player_stats))
        }
      }

      cat("⚠️  Player", name, "not found. Falling back to League Average.\n")
    }
  }

  # Default: Calculate League Average for most recent year
  recent_year <- max(Batting$yearID)
  league_avg <- Batting %>%
    filter(yearID == recent_year) %>%
    summarise(
      avg = sum(H, na.rm = TRUE) / sum(AB, na.rm = TRUE),
      obp = sum(H + BB + HBP, na.rm = TRUE) / sum(AB + BB + HBP + SF, na.rm = TRUE),
      slg = sum(H + X2B + 2*X3B + 3*HR, na.rm = TRUE) / sum(AB, na.rm = TRUE),
      pa = sum(AB + BB + HBP + SF, na.rm = TRUE)
    )

  cat("📈 Using LEAGUE AVERAGE stats for batter (", recent_year, " season)\n", sep = "")
  return(as.list(league_avg))
}

get_pitcher_stats <- function(name = NULL) {
  # Decision tree for data source selection:
  # 1. If manual_pitcher_stats exists -> Use manual stats
  # 2. If name is provided -> Fetch from Lahman database
  # 3. If name is NULL or not found -> Calculate League Average for most recent year

  # Check for manual stats first
  if (exists("manual_pitcher_stats") && !is.null(manual_pitcher_stats)) {
    cat("📊 Using MANUAL stats for pitcher\n")
    return(manual_pitcher_stats)
  }

  # If a specific name is provided, fetch from Lahman
  if (!is.null(name)) {
    # Parse name (format: "Last, First")
    name_parts <- str_split(name, ", ")[[1]]
    if (length(name_parts) == 2) {
      last_name <- name_parts[1]
      first_name <- name_parts[2]

      # Find player in People table
      player <- People %>%
        filter(nameLast == last_name, nameFirst == first_name) %>%
        slice(1)

      if (nrow(player) > 0) {
        player_id <- player$playerID

        # Get most recent season stats
        recent_year <- max(Pitching$yearID)
        player_stats <- Pitching %>%
          filter(playerID == player_id, yearID == recent_year) %>%
          summarise(
            era = sum(ER, na.rm = TRUE) * 9 / sum(IPouts, na.rm = TRUE) * 3,
            whip = sum(BB + H, na.rm = TRUE) / (sum(IPouts, na.rm = TRUE) / 3),
            so_per_9 = sum(SO, na.rm = TRUE) * 9 / (sum(IPouts, na.rm = TRUE) / 3),
            bb_per_9 = sum(BB, na.rm = TRUE) * 9 / (sum(IPouts, na.rm = TRUE) / 3),
            ip = sum(IPouts, na.rm = TRUE) / 3
          )

        if (nrow(player_stats) > 0 && player_stats$ip > 0) {
          cat("⚾ Using LAHMAN DATABASE stats for pitcher:", name, "\n")
          return(as.list(player_stats))
        }
      }

      cat("⚠️  Player", name, "not found. Falling back to League Average.\n")
    }
  }

  # Default: Calculate League Average for most recent year
  recent_year <- max(Pitching$yearID)
  league_avg <- Pitching %>%
    filter(yearID == recent_year) %>%
    summarise(
      era = sum(ER, na.rm = TRUE) * 9 / sum(IPouts, na.rm = TRUE) * 3,
      whip = sum(BB + H, na.rm = TRUE) / (sum(IPouts, na.rm = TRUE) / 3),
      so_per_9 = sum(SO, na.rm = TRUE) * 9 / (sum(IPouts, na.rm = TRUE) / 3),
      bb_per_9 = sum(BB, na.rm = TRUE) * 9 / (sum(IPouts, na.rm = TRUE) / 3),
      ip = sum(IPouts, na.rm = TRUE) / 3
    )

  cat("📈 Using LEAGUE AVERAGE stats for pitcher (", recent_year, " season)\n", sep = "")
  return(as.list(league_avg))
}

# ============================================================================
# STATCAST DATA RETRIEVAL FUNCTIONS
# ============================================================================

get_statcast_batter_stats <- function(name = NULL) {
  # Retrieve batter stats from Statcast CSV

  if (!exists("statcast_batters")) {
    stop("Statcast data not loaded. Set data_source = 'statcast' in user inputs.")
  }

  # Check for manual stats first
  if (exists("manual_batter_stats") && !is.null(manual_batter_stats)) {
    cat("📊 Using MANUAL stats for batter\n")
    return(manual_batter_stats)
  }

  # If a specific name is provided, fetch from Statcast
  if (!is.null(name)) {
    player_data <- statcast_batters %>%
      filter(player_name == name)

    if (nrow(player_data) > 0) {
      stats <- list(
        player_name = player_data$player_name[1],
        pa = player_data$pa[1],
        avg = player_data$batting_avg[1],
        obp = player_data$on_base_percent[1],
        slg = player_data$slg_percent[1],
        woba = player_data$woba[1],
        swing_pct = player_data$swing_percent[1] / 100,
        z_swing_pct = player_data$z_swing_percent[1] / 100,
        oz_swing_pct = player_data$oz_swing_percent[1] / 100,
        iz_contact_pct = player_data$iz_contact_percent[1] / 100,
        oz_contact_pct = player_data$oz_contact_percent[1] / 100,
        f_strike_pct = player_data$f_strike_percent[1] / 100
      )
      cat("⚾ Using STATCAST data for batter:", name, "\n")
      return(stats)
    } else {
      cat("⚠️  Player", name, "not found in Statcast. Falling back to League Average.\n")
    }
  }

  # Default: Calculate League Average from Statcast
  league_avg <- statcast_batters %>%
    summarise(
      avg = mean(batting_avg, na.rm = TRUE),
      obp = mean(on_base_percent, na.rm = TRUE),
      slg = mean(slg_percent, na.rm = TRUE),
      woba = mean(woba, na.rm = TRUE),
      swing_pct = mean(swing_percent, na.rm = TRUE) / 100,
      z_swing_pct = mean(z_swing_percent, na.rm = TRUE) / 100,
      oz_swing_pct = mean(oz_swing_percent, na.rm = TRUE) / 100,
      iz_contact_pct = mean(iz_contact_percent, na.rm = TRUE) / 100,
      oz_contact_pct = mean(oz_contact_percent, na.rm = TRUE) / 100,
      f_strike_pct = mean(f_strike_percent, na.rm = TRUE) / 100,
      pa = sum(pa, na.rm = TRUE)
    )

  cat("📈 Using STATCAST LEAGUE AVERAGE stats for batter (2024 season)\n")
  return(as.list(league_avg))
}

get_statcast_pitcher_stats <- function(name = NULL) {
  # Retrieve pitcher stats from Statcast CSV

  if (!exists("statcast_pitchers")) {
    stop("Statcast data not loaded. Set data_source = 'statcast' in user inputs.")
  }

  # Check for manual stats first
  if (exists("manual_pitcher_stats") && !is.null(manual_pitcher_stats)) {
    cat("📊 Using MANUAL stats for pitcher\n")
    return(manual_pitcher_stats)
  }

  # If a specific name is provided, fetch from Statcast
  if (!is.null(name)) {
    player_data <- statcast_pitchers %>%
      filter(player_name == name)

    if (nrow(player_data) > 0) {
      stats <- list(
        player_name = player_data$player_name[1],
        pa = player_data$pa[1],
        woba = player_data$woba[1],
        xwoba = player_data$xwoba[1],
        f_strike_pct = player_data$f_strike_percent[1] / 100,
        whiff_pct = player_data$whiff_percent[1] / 100,
        z_swing_pct = player_data$z_swing_percent[1] / 100,
        oz_swing_pct = player_data$oz_swing_percent[1] / 100,
        swing_pct = player_data$swing_percent[1] / 100,
        hard_hit_pct = player_data$hard_hit_percent[1] / 100
      )
      cat("⚾ Using STATCAST data for pitcher:", name, "\n")
      return(stats)
    } else {
      cat("⚠️  Player", name, "not found in Statcast. Falling back to League Average.\n")
    }
  }

  # Default: Calculate League Average from Statcast
  league_avg <- statcast_pitchers %>%
    summarise(
      woba = mean(woba, na.rm = TRUE),
      xwoba = mean(xwoba, na.rm = TRUE),
      f_strike_pct = mean(f_strike_percent, na.rm = TRUE) / 100,
      whiff_pct = mean(whiff_percent, na.rm = TRUE) / 100,
      z_swing_pct = mean(z_swing_percent, na.rm = TRUE) / 100,
      oz_swing_pct = mean(oz_swing_percent, na.rm = TRUE) / 100,
      swing_pct = mean(swing_percent, na.rm = TRUE) / 100,
      hard_hit_pct = mean(hard_hit_percent, na.rm = TRUE) / 100,
      pa = sum(pa, na.rm = TRUE)
    )

  cat("📈 Using STATCAST LEAGUE AVERAGE stats for pitcher (2024 season)\n")
  return(as.list(league_avg))
}
```

```{r load_data}
# ============================================================================
# LOAD DATA FOR ANALYSIS
# ============================================================================

# Retrieve batter and pitcher statistics based on data source selection
if (data_source == "statcast") {
  batter_stats <- get_statcast_batter_stats(target_batter)
  pitcher_stats <- get_statcast_pitcher_stats(target_pitcher)
} else {
  batter_stats <- get_batter_stats(target_batter)
  pitcher_stats <- get_pitcher_stats(target_pitcher)
}
```

## Game Theory Matrix

```{r payoff_matrix}
# ============================================================================
# 2x2 PAYOFF MATRIX: First Pitch Game
# ============================================================================

# Run Expectancy Deltas (from batter's perspective)
# These values represent the change in expected runs for the inning
# based on count transitions from 0-0

# Count transitions:
# 0-0 to 0-1 (strike): -0.06 runs
# 0-0 to 1-0 (ball):   +0.04 runs

re_delta_strike <- -0.06
re_delta_ball <- 0.04

# Zone + Swing outcome: Based on contact quality
# Using league average OBP as proxy for successful contact value
# Typical value range: +0.15 to +0.30 runs
re_delta_contact <- batter_stats$obp * 0.60  # Scaled contact outcome

# Waste + Swing outcome: Chase/whiff leads to 0-1 count
# Same as called strike but with risk of whiff
re_delta_chase <- re_delta_strike

# Build the 2x2 payoff matrix
# Rows: Pitcher's strategy (Zone vs Waste)
# Columns: Batter's strategy (Swing vs Take)
payoff_matrix <- tibble(
  pitcher_strategy = c("Zone", "Waste"),
  swing = c(re_delta_contact, re_delta_chase),
  take = c(re_delta_strike, re_delta_ball)
)

# Display the matrix
cat("\n🎯 First Pitch Payoff Matrix (Batter's Expected Runs)\n")
cat("=" , rep("=", 50), "\n", sep = "")
print(payoff_matrix, width = Inf)
cat("\n")
cat("Legend:\n")
cat("  Zone + Swing = Contact outcome (", sprintf("%.3f", re_delta_contact), " runs)\n", sep = "")
cat("  Zone + Take  = Called strike (", sprintf("%.3f", re_delta_strike), " runs)\n", sep = "")
cat("  Waste + Swing = Chase/whiff   (", sprintf("%.3f", re_delta_chase), " runs)\n", sep = "")
cat("  Waste + Take = Ball           (", sprintf("%.3f", re_delta_ball), " runs)\n", sep = "")
```

## Nash Equilibrium Solver

```{r nash_equilibrium}
# ============================================================================
# MIXED STRATEGY NASH EQUILIBRIUM SOLVER
# ============================================================================

solve_nash_equilibrium <- function(matrix) {
  # Extract payoff values from the matrix
  # Matrix structure (batter's perspective):
  #              Swing       Take
  # Zone     [contact]   [strike]
  # Waste    [chase]     [ball]

  zone_swing <- matrix$swing[matrix$pitcher_strategy == "Zone"]
  zone_take <- matrix$take[matrix$pitcher_strategy == "Zone"]
  waste_swing <- matrix$swing[matrix$pitcher_strategy == "Waste"]
  waste_take <- matrix$take[matrix$pitcher_strategy == "Waste"]

  # Calculate batter's optimal swing probability (p)
  # Batter is indifferent when EV(Swing) = EV(Take) across pitcher's strategies
  # Solving: p*(zone_swing - waste_swing) + (1-p)*(zone_take - waste_take) makes pitcher indifferent
  # Rearranged: p = (waste_take - zone_take) / (zone_swing - waste_swing + waste_take - zone_take)

  numerator_batter <- waste_take - zone_take
  denominator_batter <- zone_swing - waste_swing + waste_take - zone_take

  optimal_swing_pct <- numerator_batter / denominator_batter

  # Calculate pitcher's optimal zone probability (q)
  # Pitcher is indifferent when EV(Zone) = EV(Waste) across batter's strategies
  # Solving: q = (waste_take - waste_swing) / (zone_swing - zone_take + waste_take - waste_swing)

  numerator_pitcher <- waste_take - waste_swing
  denominator_pitcher <- zone_swing - zone_take + waste_take - waste_swing

  optimal_zone_pct <- numerator_pitcher / denominator_pitcher

  # Calculate equilibrium expected value for batter
  eq_value <- optimal_zone_pct * zone_swing + (1 - optimal_zone_pct) * waste_swing

  # Return results as a list
  list(
    batter_swing_pct = optimal_swing_pct,
    pitcher_zone_pct = optimal_zone_pct,
    equilibrium_value = eq_value,
    payoffs = list(
      zone_swing = zone_swing,
      zone_take = zone_take,
      waste_swing = waste_swing,
      waste_take = waste_take
    )
  )
}

# Solve for Nash equilibrium
equilibrium <- solve_nash_equilibrium(payoff_matrix)

# Display results
cat("\n⚖️  NASH EQUILIBRIUM SOLUTION\n")
cat("=" , rep("=", 50), "\n", sep = "")
cat("\n🏏 BATTER'S OPTIMAL STRATEGY:\n")
cat("   Swing on first pitch: ", sprintf("%.1f%%", equilibrium$batter_swing_pct * 100), "\n", sep = "")
cat("   Take on first pitch:  ", sprintf("%.1f%%", (1 - equilibrium$batter_swing_pct) * 100), "\n", sep = "")
cat("\n⚾ PITCHER'S OPTIMAL STRATEGY:\n")
cat("   Throw in zone:  ", sprintf("%.1f%%", equilibrium$pitcher_zone_pct * 100), "\n", sep = "")
cat("   Throw waste:    ", sprintf("%.1f%%", (1 - equilibrium$pitcher_zone_pct) * 100), "\n", sep = "")
cat("\n📊 EQUILIBRIUM EXPECTED VALUE:\n")
cat("   Batter's expected runs: ", sprintf("%.4f", equilibrium$equilibrium_value), "\n", sep = "")
cat("\n")
```

## Best Response to Real Data

```{r best_response}
# ============================================================================
# BEST RESPONSE ANALYSIS: Using Real Pitcher Behavior
# ============================================================================

# This section calculates the batter's optimal swing % given the ACTUAL
# first-pitch strike rate observed in real data (not theoretical equilibrium)

if (data_source == "statcast" && !is.null(pitcher_stats$f_strike_pct)) {

  # Extract observed pitcher zone rate from Statcast data
  observed_zone_pct <- pitcher_stats$f_strike_pct

  # Calculate batter's expected value for any swing percentage
  # given the FIXED observed pitcher behavior
  calc_batter_ev <- function(swing_pct, zone_pct) {
    zone_swing <- equilibrium$payoffs$zone_swing
    zone_take <- equilibrium$payoffs$zone_take
    waste_swing <- equilibrium$payoffs$waste_swing
    waste_take <- equilibrium$payoffs$waste_take

    # Expected value = (prob zone) * EV(zone) + (prob waste) * EV(waste)
    ev <- zone_pct * (swing_pct * zone_swing + (1 - swing_pct) * zone_take) +
          (1 - zone_pct) * (swing_pct * waste_swing + (1 - swing_pct) * waste_take)
    return(ev)
  }

  # Find optimal swing percentage by testing all values
  swing_range <- seq(0, 1, by = 0.001)
  ev_values <- sapply(swing_range, function(p) calc_batter_ev(p, observed_zone_pct))
  optimal_idx <- which.max(ev_values)
  optimal_swing_pct_real <- swing_range[optimal_idx]
  optimal_ev_real <- ev_values[optimal_idx]

  # Display results
  cat("\n🎯 BEST RESPONSE TO REAL PITCHER DATA\n")
  cat("=" , rep("=", 50), "\n", sep = "")
  cat("\n⚾ OBSERVED PITCHER BEHAVIOR:\n")
  cat("   First-pitch strike rate: ", sprintf("%.1f%%", observed_zone_pct * 100), "\n", sep = "")
  cat("\n🏏 BATTER'S OPTIMAL RESPONSE:\n")
  cat("   Swing on first pitch: ", sprintf("%.1f%%", optimal_swing_pct_real * 100), "\n", sep = "")
  cat("   Take on first pitch:  ", sprintf("%.1f%%", (1 - optimal_swing_pct_real) * 100), "\n", sep = "")
  cat("\n📊 EXPECTED VALUE AT OPTIMAL:\n")
  cat("   Batter's expected runs: ", sprintf("%.4f", optimal_ev_real), "\n", sep = "")
  cat("\n💡 COMPARISON:\n")
  cat("   Nash Equilibrium swing%:  ", sprintf("%.1f%%", equilibrium$batter_swing_pct * 100), "\n", sep = "")
  cat("   Best Response swing%:     ", sprintf("%.1f%%", optimal_swing_pct_real * 100), "\n", sep = "")

  if (observed_zone_pct > equilibrium$pitcher_zone_pct) {
    cat("   → Pitcher throws MORE strikes than equilibrium\n")
    cat("   → Batter should be MORE aggressive\n")
  } else if (observed_zone_pct < equilibrium$pitcher_zone_pct) {
    cat("   → Pitcher throws FEWER strikes than equilibrium\n")
    cat("   → Batter should be MORE patient\n")
  } else {
    cat("   → Pitcher behavior matches equilibrium perfectly\n")
  }
  cat("\n")

} else {
  cat("\n📚 Best Response analysis requires Statcast data\n")
  cat("   Set data_source = 'statcast' to enable this feature\n\n")
}
```

## Visualization

```{r equilibrium_plot, fig.width=10, fig.height=6}
# ============================================================================
# EXPECTED VALUE CURVES: Visualizing Nash Equilibrium
# ============================================================================

# Create sequence of swing probabilities (0% to 100%)
swing_prob_seq <- seq(0, 1, by = 0.01)

# Calculate expected values for batter across different swing probabilities
# For a given swing probability p:
# EV(Zone) = p * zone_swing + (1-p) * zone_take
# EV(Waste) = p * waste_swing + (1-p) * waste_take

ev_data <- tibble(
  swing_pct = swing_prob_seq * 100,
  ev_if_zone = swing_prob_seq * equilibrium$payoffs$zone_swing +
               (1 - swing_prob_seq) * equilibrium$payoffs$zone_take,
  ev_if_waste = swing_prob_seq * equilibrium$payoffs$waste_swing +
                (1 - swing_prob_seq) * equilibrium$payoffs$waste_take
) %>%
  pivot_longer(
    cols = c(ev_if_zone, ev_if_waste),
    names_to = "pitcher_location",
    values_to = "expected_value"
  ) %>%
  mutate(
    pitcher_location = case_when(
      pitcher_location == "ev_if_zone" ~ "Pitcher Throws Zone",
      pitcher_location == "ev_if_waste" ~ "Pitcher Throws Waste"
    )
  )

# Create the plot
ggplot(ev_data, aes(x = swing_pct, y = expected_value, color = pitcher_location)) +
  geom_line(linewidth = 1.2) +
  geom_vline(
    xintercept = equilibrium$batter_swing_pct * 100,
    linetype = "dashed",
    color = baseball_colors$red,
    linewidth = 0.8
  ) +
  geom_point(
    data = tibble(
      x = equilibrium$batter_swing_pct * 100,
      y = equilibrium$equilibrium_value
    ),
    aes(x = x, y = y),
    color = baseball_colors$red,
    size = 4,
    inherit.aes = FALSE
  ) +
  annotate(
    "text",
    x = equilibrium$batter_swing_pct * 100,
    y = equilibrium$equilibrium_value + 0.02,
    label = sprintf("Nash Equilibrium\nSwing: %.1f%%", equilibrium$batter_swing_pct * 100),
    color = baseball_colors$red,
    fontface = "bold",
    size = 4,
    vjust = 0
  ) +
  scale_color_manual(
    values = c("Pitcher Throws Zone" = baseball_colors$dark_green,
               "Pitcher Throws Waste" = baseball_colors$blue)
  ) +
  labs(
    title = "First Pitch Expected Value: Finding Nash Equilibrium",
    subtitle = "Batter's optimal swing percentage makes pitcher indifferent between Zone and Waste",
    x = "Batter Swing Percentage (%)",
    y = "Expected Runs (Batter's Perspective)",
    color = "Pitcher Strategy"
  ) +
  theme_baseball() +
  scale_x_continuous(breaks = seq(0, 100, 20), limits = c(0, 100)) +
  scale_y_continuous(labels = label_number(accuracy = 0.01))
```

```{r comparison_plot, fig.width=10, fig.height=6}
# ============================================================================
# COMPARISON VISUALIZATION: Nash vs. Best Response
# ============================================================================

if (data_source == "statcast" && exists("optimal_swing_pct_real")) {

  # Create data for visualization
  swing_seq <- seq(0, 1, by = 0.01)

  # Calculate EV for observed pitcher behavior across swing percentages
  ev_observed <- sapply(swing_seq, function(p) calc_batter_ev(p, observed_zone_pct))

  comparison_data <- tibble(
    swing_pct = swing_seq * 100,
    ev = ev_observed
  )

  # Create the plot
  ggplot(comparison_data, aes(x = swing_pct, y = ev)) +
    geom_line(linewidth = 1.2, color = baseball_colors$dark_green) +

    # Nash Equilibrium point
    geom_vline(
      xintercept = equilibrium$batter_swing_pct * 100,
      linetype = "dashed",
      color = baseball_colors$blue,
      linewidth = 0.8
    ) +
    geom_point(
      data = tibble(
        x = equilibrium$batter_swing_pct * 100,
        y = calc_batter_ev(equilibrium$batter_swing_pct, observed_zone_pct)
      ),
      aes(x = x, y = y),
      color = baseball_colors$blue,
      size = 4
    ) +
    annotate(
      "text",
      x = equilibrium$batter_swing_pct * 100,
      y = calc_batter_ev(equilibrium$batter_swing_pct, observed_zone_pct) - 0.015,
      label = sprintf("Nash Equilibrium\n%.1f%%", equilibrium$batter_swing_pct * 100),
      color = baseball_colors$blue,
      fontface = "bold",
      size = 3.5,
      vjust = 1
    ) +

    # Best Response point
    geom_vline(
      xintercept = optimal_swing_pct_real * 100,
      linetype = "solid",
      color = baseball_colors$red,
      linewidth = 0.8
    ) +
    geom_point(
      data = tibble(
        x = optimal_swing_pct_real * 100,
        y = optimal_ev_real
      ),
      aes(x = x, y = y),
      color = baseball_colors$red,
      size = 4
    ) +
    annotate(
      "text",
      x = optimal_swing_pct_real * 100,
      y = optimal_ev_real + 0.015,
      label = sprintf("Best Response\n%.1f%%", optimal_swing_pct_real * 100),
      color = baseball_colors$red,
      fontface = "bold",
      size = 3.5,
      vjust = 0
    ) +

    labs(
      title = "Optimal Swing% vs. Real Pitcher Behavior",
      subtitle = sprintf("Given observed first-pitch strike rate: %.1f%%", observed_zone_pct * 100),
      x = "Batter Swing Percentage (%)",
      y = "Expected Runs (Batter's Perspective)"
    ) +
    theme_baseball() +
    scale_x_continuous(breaks = seq(0, 100, 20), limits = c(0, 100)) +
    scale_y_continuous(labels = label_number(accuracy = 0.01))
}
```

## Strategic Interpretation

```{r interpretation}
# ============================================================================
# AUTOMATIC INTERPRETATION OF RESULTS
# ============================================================================

cat("\n📖 STRATEGIC ANALYSIS\n")
cat("=" , rep("=", 70), "\n", sep = "")
cat("\n")

# Determine player types based on current configuration
if (!is.null(target_batter)) {
  batter_type <- paste0("batter ", target_batter)
} else {
  batter_type <- "the league average batter"
}

if (!is.null(target_pitcher)) {
  pitcher_type <- paste0("pitcher ", target_pitcher)
} else {
  pitcher_type <- "the league average pitcher"
}

cat("MATCHUP: ", batter_type, " vs. ", pitcher_type, "\n\n", sep = "")

cat("EQUILIBRIUM STRATEGY:\n")
cat("  • The batter should swing on ", sprintf("%.1f%%", equilibrium$batter_swing_pct * 100),
    " of first pitches\n", sep = "")
cat("  • The pitcher should throw in the zone ", sprintf("%.1f%%", equilibrium$pitcher_zone_pct * 100),
    " of the time\n\n", sep = "")

cat("WHY THIS WORKS:\n")
cat("  At this equilibrium, both players are mixing their strategies optimally.\n")
cat("  Neither player can improve their expected outcome by changing strategy\n")
cat("  unilaterally, which defines the Nash Equilibrium.\n\n")

# Assess aggressiveness
if (equilibrium$batter_swing_pct > 0.50) {
  swing_assessment <- "aggressive"
  swing_context <- "This batter should be looking to attack first pitches."
} else if (equilibrium$batter_swing_pct < 0.35) {
  swing_assessment <- "patient"
  swing_context <- "This batter should be selective on first pitches."
} else {
  swing_assessment <- "balanced"
  swing_context <- "This batter should use a moderate approach on first pitches."
}

cat("PRACTICAL IMPLICATIONS:\n")
cat("  The optimal swing rate of ", sprintf("%.1f%%", equilibrium$batter_swing_pct * 100),
    " indicates a ", swing_assessment, " approach.\n", sep = "")
cat("  ", swing_context, "\n\n", sep = "")

cat("EXPECTED OUTCOME:\n")
cat("  At equilibrium, the batter's expected value is ",
    sprintf("%.4f", equilibrium$equilibrium_value), " runs.\n", sep = "")

if (equilibrium$equilibrium_value > 0) {
  cat("  This positive value favors the batter in this matchup.\n")
} else if (equilibrium$equilibrium_value < 0) {
  cat("  This negative value favors the pitcher in this matchup.\n")
} else {
  cat("  This neutral value indicates an evenly matched situation.\n")
}

cat("\n")

# Add Statcast-specific insights
if (data_source == "statcast" && exists("optimal_swing_pct_real")) {
  cat("=" , rep("=", 70), "\n", sep = "")
  cat("\n📊 STATCAST DATA INSIGHTS\n")
  cat("=" , rep("=", 70), "\n", sep = "")
  cat("\n")

  cat("REAL-WORLD DATA:\n")
  cat("  • Observed first-pitch strike rate: ", sprintf("%.1f%%", observed_zone_pct * 100), "\n", sep = "")
  cat("  • Nash equilibrium predicts: ", sprintf("%.1f%%", equilibrium$pitcher_zone_pct * 100), "\n", sep = "")

  deviation <- abs(observed_zone_pct - equilibrium$pitcher_zone_pct)
  if (deviation > 0.10) {
    cat("  • LARGE deviation from equilibrium (", sprintf("%.1f", deviation * 100), " percentage points)\n\n", sep = "")
  } else if (deviation > 0.05) {
    cat("  • Moderate deviation from equilibrium (", sprintf("%.1f", deviation * 100), " percentage points)\n\n", sep = "")
  } else {
    cat("  • Close to equilibrium prediction (", sprintf("%.1f", deviation * 100), " percentage points)\n\n", sep = "")
  }

  cat("ACTIONABLE STRATEGY:\n")

  if (observed_zone_pct > equilibrium$pitcher_zone_pct + 0.05) {
    cat("  ⚠️  PITCHER THROWS TOO MANY STRIKES\n")
    cat("  → Batter can exploit this by being MORE aggressive\n")
    cat("  → Recommended swing%: ", sprintf("%.1f%%", optimal_swing_pct_real * 100), "\n", sep = "")
    cat("  → This is ", sprintf("%.1f", (optimal_swing_pct_real - equilibrium$batter_swing_pct) * 100),
        " percentage points MORE aggressive than equilibrium\n", sep = "")
    ev_gain <- optimal_ev_real - equilibrium$equilibrium_value
    if (ev_gain > 0) {
      cat("  → Expected value gain: +", sprintf("%.4f", ev_gain), " runs per first pitch\n", sep = "")
    }
  } else if (observed_zone_pct < equilibrium$pitcher_zone_pct - 0.05) {
    cat("  ⚠️  PITCHER THROWS TOO FEW STRIKES\n")
    cat("  → Batter should be MORE patient and wait for better counts\n")
    cat("  → Recommended swing%: ", sprintf("%.1f%%", optimal_swing_pct_real * 100), "\n", sep = "")
    cat("  → This is ", sprintf("%.1f", abs(optimal_swing_pct_real - equilibrium$batter_swing_pct) * 100),
        " percentage points LESS aggressive than equilibrium\n", sep = "")
    ev_gain <- optimal_ev_real - equilibrium$equilibrium_value
    if (ev_gain > 0) {
      cat("  → Expected value gain: +", sprintf("%.4f", ev_gain), " runs per first pitch\n", sep = "")
    }
  } else {
    cat("  ✅ PITCHER BEHAVIOR NEAR EQUILIBRIUM\n")
    cat("  → Nash equilibrium strategy is approximately optimal\n")
    cat("  → Recommended swing%: ", sprintf("%.1f%%", optimal_swing_pct_real * 100), "\n", sep = "")
  }

  cat("\n")
}

cat("=" , rep("=", 70), "\n", sep = "")
```

---

**Analysis Complete.**

**To explore different scenarios:**
- Change `data_source` to switch between "lahman" and "statcast"
- Adjust `target_batter` and `target_pitcher` to analyze specific matchups
- Set both to `NULL` for league average comparisons
- Use Statcast data to see how real pitcher behavior differs from theoretical equilibrium


---

**Analysis Complete.** Adjust `target_batter` and `target_pitcher` in the User Configuration section to explore different matchups.
